---
title: EmbedSOM vs. archeology
author: Miroslav Kratochvíl
email: exa.exa@gmail.com
date: 2020-01-18
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align="center", fig.height=6, fig.width=7, cache.extra = knitr::rand_seed)
```

Smithsonian Institute provides a whole load of interesting data, among them a 3D model of a [woolly mammoth skeleton](https://3d.si.edu/explorer/woolly-mammoth) and of a [T-rex skeleton eating triceratops skeleton](https://3d.si.edu/t-rex)!

In this vignette, we convert them to flat versions.

# Getting the data

The data is available as STL models. You should be able to get a list of 3D point coordinates from the STL either directly, or using some of the available commandline tools (e.g. `stl2gts`). At the end, you should end up having a 3-column matrix of point coordinates. I saved them to `mammoth.points` and `trex.points` and loaded them accordingly. The models have hundreds of thousands of individual points!

```{r}
mammoth <- read.table('mammoth.points', header=F)
trex <- read.table('trex.points', header=F)
print(dim(mammoth))
print(dim(trex))
```

The data is organized as expected:

```{r}
par(mar=rep(0,4))
EmbedSOM::PlotEmbed(mammoth[,c(2,3)])
```
```{r}
par(mar=rep(0,4))
EmbedSOM::PlotEmbed(-trex[,c(1,2)])
```

# A flat woolly mammoth

We may embed the mammoth using the "standard" approach. I use an extra large grid to get extra detail, but smaller grids usually suffice (and may compute much faster).

```{r}
set.seed(1)
print(system.time(
e <- EmbedSOM::EmbedSOM(mammoth, parallel=T,
       map=EmbedSOM::SOM(mammoth, xdim=32, ydim=32, parallel=T, batch=T))
)[3])
```

Let us color the mammoth components by color, so that we see which leg belongs where:

```{r}
par(mar=rep(0,4))
EmbedSOM::PlotEmbed(e, data=mammoth, red=1, green=2, blue=3, alpha=.1)
```

We can observe several things:

- Although the SOM has decided to "cut" the mammoth chest open in order to picture it nicely, it did not cut the ribs, just stretched them quite a bit.
- There is apparently a lot of pressure for putting the legs to one side of the picture, which we might want to reduce.
- Tusks are rendered pretty tiny; mostly because they are underrepresented in the dataset.

Let's try to fix some of the problems:
```{r}
set.seed(1)
mammoth[,3]<-mammoth[,3]*0.5 #pretend the mammoth is not that tall
print(system.time(
e <- EmbedSOM::EmbedSOM(mammoth, parallel=T,
       map=EmbedSOM::SOM(mammoth, xdim=24, ydim=24, rlen=20, parallel=T, batch=T))
)[3])
par(mar=rep(0,4))
EmbedSOM::PlotEmbed(e, data=mammoth, red=1, green=2, blue=3, alpha=.1)
```

Almost good.

# A flat T-rex eating a flat triceratops

The breakfast scene is slightly overcrowded if embedded by plain SOMs:

```{r}
set.seed(1)
print(system.time(
e <- EmbedSOM::EmbedSOM(trex, parallel=T,
       map=EmbedSOM::SOM(trex, xdim=32, ydim=32, rlen=20, parallel=T, batch=T))
)[3])
par(mar=rep(0,4))
EmbedSOM::PlotEmbed(e, data=trex, blue=3, alpha=.2)
```

SOM-less embedding (with random landmarks) alleviates this problem (we use UMAP from the `uwot` package to organize 1000 randomly chosen landmarks, and project the rest of the dataset there):

```{r}
set.seed(1)
print(system.time(
e <- EmbedSOM::EmbedSOM(trex, parallel=T,
       map=EmbedSOM::RandomMap(trex, 1000, coordsFn=EmbedSOM::uwotCoords(min_dist=2)))
)[3])
par(mar=rep(0,4))
EmbedSOM::PlotEmbed(e, data=trex, blue=3, alpha=.2)
```
